Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-05 paper/main.tex
Martin's suggestins reflected

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parameterized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G} to get generic TELA) from the Spot
library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA by
\texttt{ltl3tela} with option \texttt{-D1} and by two state-of-the-art
translators from the Owl library~\cite{kretinsky.18.atva}, namely
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} and
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav}. Some
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator for one of the two formula
sets), we have applied all single-level reductions and the multi-level
reduction, always with incremental approach and without reducing the
number of cubes. We do not reduce the number of cubes as our primary
aim is to reduce the number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

@@

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata set and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. One can observe that a higher level sometimes achieves a smaller
reduction than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl3tela} on automata coming from formulas from
literature). This is caused by the QBF solver timeouts occurring
earlier as formulas constructed by the higher level are more
complex. The automata sets produced by \texttt{ltl2dela} and
\texttt{ltl2dgra}, on formulas from literature do not contain any
automaton where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

Table~\ref{tab:heatmap} shows the number of automata from all five
sets constructed from formulas from literature that have a given
number of original acceptance marks and a given number of acceptance
marks after multi-level reduction. The table indicates that in many
cases only 1 or 2~marks can be saved. However, the achieved reduction
is substantial for some automata with a higher number of original
acceptance marks. For example, in 26 cases, we have reduced 7 or more
acceptance marks to only 4 or less.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.
Table~\ref{tab:heatmap-random} shows the same information for all
considered automata constructed from random formulas.

@@

\begin{figure}[p]
  \centering
  \input{quantile.tex}
  \bigskip\bigskip
  \input{quantile-random.tex}
  \bigskip
  \caption{Performance of \telatko on individual automata of each automata set.
    Automata sets constructed from formulas from literature are in the upper graph,
    automata sets constructed from random formulas are in the lower graph.
    Each line shows the time ($y$ axis) needed by \telatko to process the
    $x^\textrm{th}$ automaton of the set, where automata in the set are ordered by
    their processing time.
    % The $x$ axis represents the number of automata from each automata
    % set that are processed by \telatko in the time shown on the $y$
    % axis.
  }
\label{fig:quantile}\end{figure}

>>>>

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parameterized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348~LTL
formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G} to get generic TELA) from the Spot
library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA by
\texttt{ltl3tela} with option \texttt{-D1} and by two state-of-the-art
translators from the Owl library~\cite{kretinsky.18.atva}, namely
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} and
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav}. Some
translators failed on some formulas: they usually reached a timeout of
60 seconds or produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

To all automata,
% To each automata set (that is, the automata with at least two marks
% produced by an individual LTL translator for one of the two formula
% sets),
we have applied all single-level reductions and the multi-level
reduction, always with incremental approach and without reducing the
number of cubes. We do not reduce the number of cubes as our primary
aim is to reduce the number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32~GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

@@

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata set and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. One can observe that a higher level sometimes achieves a smaller
reduction than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl3tela} on automata coming from formulas from
literature). This is caused by the QBF solver timeouts occurring
earlier as formulas constructed by the higher level are more
complex. The automata sets produced by \texttt{ltl2dela} and
\texttt{ltl2dgra} on formulas from literature do not contain any
automaton where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

Table~\ref{tab:heatmap} shows the number of all automata constructed
from formulas from literature that have a given number of original
acceptance marks and a given number of acceptance marks after
multi-level reduction. The table indicates that in many cases only 1
or 2~marks can be saved. However, the achieved reduction is
substantial for some automata with a higher number of original
acceptance marks. For example, in 26 cases, we have reduced 7 or more
acceptance marks to only 4 or less.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.
Table~\ref{tab:heatmap-random} shows the same information for all
automata constructed from random formulas.

@@

\begin{figure}[p]
  \centering
  \input{quantile.tex}
  \bigskip\bigskip
  \input{quantile-random.tex}
  \bigskip
  \caption{Running times of \telatko on individual automata of each automata set.
    Automata sets constructed from formulas from literature are in the upper graph,
    automata sets constructed from random formulas are in the lower graph.
    Each line shows the time ($y$ axis) needed by \telatko to process the
    $x^\textrm{th}$ automaton of the set, where automata in the set are ordered by
    their processing time.
    % The $x$ axis represents the number of automata from each automata
    % set that are processed by \telatko in the time shown on the $y$
    % axis.
  }
\label{fig:quantile}\end{figure}
